---
title: "Introduction to pactR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to pactR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

**pactR** is an R interface to *PACT* — **P**osterior **A**nalysis of **C**oalescent **T**rees
(Bedford 2009–2013). It reads posterior samples of labeled phylogenetic trees produced by
programs such as BEAST, Migrate, IM, and LAMARC, and computes structured coalescent
statistics including:

- TMRCA and total branch length
- Per-population coalescent rates
- Directional migration rates
- Nucleotide diversity, F~ST~, and Tajima's D
- Spatial diffusion coefficients and drift rates
- Skyline (time-windowed) versions of all statistics
- Per-tip time-to-trunk and location histories

## Installation

Install the package from source:

```r
# If you have the source directory:
install.packages("path/to/pactR", repos = NULL, type = "source")
```

## Quick start

```{r example, eval=FALSE}
library(pactR)

# Get paths to bundled example files
ex <- pact_example_files()

# Run the full analysis
outdir <- tempdir()
results <- run_pact(
  trees_file = ex$trees,
  param_file = ex$param,
  output_dir = outdir
)

# Read results into R data frames
stats    <- read_pact_stats(results$stats)
tips     <- read_pact_tips(results$tips)
skylines <- read_pact_skylines(results$skylines)

head(stats)
head(tips)
```

## Input files

### Tree file (`in.trees`)

A plain-text file containing one NEWICK-format tree per line. Trees may include:

- Tip labels with population codes (e.g., `1EU514626` where the first character is the
  population label).
- BEAST-style log-probability annotations: `[&lnP=-8734.21]`
- Migrate-style log-likelihood lines: `ln(L) = -8734.21`
- Migration event annotations: `[&M 5 0:0.000428597]`

### Parameter file (`in.param`)

A plain-text file with one directive per line. Comments begin with `#`. Blank lines are
ignored. See `pact_parameter_reference()` for the complete list.

**Example parameter file:**

```
push times back 2002 2007.997   # align tree times to calendar dates
renew trunk 1                   # redefine trunk at present
print rule tree                 # output best tree in Mathematica format

summary tmrca                   # time to most recent common ancestor
summary coal rates              # per-population coalescent rates
summary mig rates               # directional migration rates

skyline settings 2002 2008 0.01 # time-windowed stats: start end step
skyline proportions             # population proportions over time

tips time to trunk              # per-tip time-to-trunk statistic
```

## Output files

| File          | Contents                                         | Read with               |
|:--------------|:-------------------------------------------------|:------------------------|
| `out.stats`   | Summary statistics (lower, mean, upper)          | `read_pact_stats()`     |
| `out.tips`    | Per-tip statistics                               | `read_pact_tips()`      |
| `out.skylines`| Time-windowed statistics                         | `read_pact_skylines()`  |
| `out.rules`   | Best tree in Mathematica rule-list format        | —                       |
| `out.pairs`   | Pairwise tip statistics                          | `read_pact_pairs()`     |

## Plotting skylines

```{r plot, eval=FALSE}
library(ggplot2)

skylines <- read_pact_skylines(results$skylines)

# Plot population proportions over time
pro <- subset(skylines, grepl("^pro_", statistic))

ggplot(pro, aes(x = time, y = mean, colour = statistic)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = statistic), alpha = 0.2) +
  labs(x = "Year", y = "Proportion", title = "Population proportions over time") +
  theme_minimal()
```

## Parameter reference

For a full listing of all available parameters:

```{r params, eval=FALSE}
pact_parameter_reference()
```

The PDF user manual (Bedford 2009–2013) can be opened with:

```{r manual, eval=FALSE}
pact_manual()
```

## Citation

If you use PACT in your research, please cite:

> Bedford T (2009–2013). *PACT: Posterior Analysis of Coalescent Trees*,
> version 0.9.4. <https://github.com/trvrb/pact>
